Reproducible Research Project 1
=========================================
by: Ryan Allen

##Setting Global Options
```{r globaloptions, echo=TRUE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

## Loading Data
```{r dataload}
## Reading in the csv file
data <- read.csv("activity.csv", header = TRUE)

## Formating the date column
data$date <- as.POSIXct(data$date, format="%Y-%m-%d")

## Making a dataframe with weekday, steps, and intervals
data <- data.frame(date=data$date,
        weekday=tolower(weekdays(data$date)), 
        steps=data$steps, 
        interval=data$interval)

## Getting a look at the dataframe
head(data)
```

## Histogram of Daily Step Counts
```{r dailystepshistogram}
## Findin the sum of steps for each day
dailysteps <- with(data, aggregate(steps, by=list(data$date), sum, na.rm=T))

## Looking at the data
head(dailysteps)

## Creating the histogram of the total number of steps taken each day
hist(dailysteps$x, xlab = "Total Daily Steps", main = "Histogram of Total Number of Steps Taken Each Day", col = "green")
```

### Mean and Median Steps by Day
```{r meanandmedian}
## The equations to find the mean and median
mean <- mean(dailysteps$x)
median <- median(dailysteps$x)
```
The mean is `r mean` and the median is `r median`.

## What is the average daily activity pattern?
```{r averagepattern}
## Taking the average of each time interval
averageData <- with(data, aggregate(steps, by=list(data$interval), mean, na.rm=T))

## Renaming the columns to make the data more manageable
names(averageData) <- c("Interval","Average.Steps")


## Time series graph
plot(averageData$Interval, averageData$Average.Steps, 
     type = "l", 
     col = "black", 
     xlab="Interval", 
     ylab = "Average Steps", 
     main= "Time Series Graph of Number of Steps by Interval")
```

### Which interval has the highest average steps?
```{r highestinterval}
## Finding which entry has the highest average and assigns the interval to max_interval
max_interval <- averageData[(which(averageData$Average.Steps == max(averageData$Average.Steps))), 1]
```
The interval with the most steps day over day is interval `r max_interval`.


## Imputing Missing Values
### Number of Missing Values
```{r numberNAs}
## Finding how many NA values there are in the steps column of our data
NAs <- sum(is.na(data$steps))
```
There are `r NAs` missing values in our data set.

### Imputing Values
```{r impute}
## In data, find values in steps column = NA and replace with mean of steps
## This devises the plan and creates a new dataframe with all values replaced
data1 <- data.frame(date=data$date, 
                   weekday=(weekdays(data$date)), 
                   steps=data$steps, 
                   interval=data$interval)


data$steps[is.na(data$steps)] = mean(data$steps, na.rm = TRUE)
```

### Histogram with missing values imputed with the mean.
```{r histogramNew}
## Findin the sum of steps for each day
dailysteps <- with(data, aggregate(steps, by=list(data$date), sum, na.rm=T))

## Looking at the data
head(dailysteps)

## Creating the histogram of the total number of steps taken each day
hist(dailysteps$x, xlab = "Total Daily Steps", main = "Histogram of Total Number of Steps Taken Each Day with Imputed Values", col = "magenta")
```
```{r mean&median}
mean1 <- as.numeric(mean(dailysteps$x))
median1 <- median(dailysteps$x)
```
The new mean is `r mean1` and the new median is `r median`. 
I would say it has a slight impact, raising both the mean and median by a couple hundred points each and it made our histogram more normallly distributed.

## Difference between weekdays and weekends
```{r daytype}
## this adds a factor column in our data set with the type of day
data1 <- cbind(data, type= ifelse(data$weekday=="Saturday" | data$weekday == "Sunday", "weekend", "weekday"))

## This finds the average steps of each interval across type and weekday
averageData1 <- aggregate(data1$steps, by=list(data1$weekday, data1$type, data1$interval), mean)

## Replaces the names in the new dataframe with the appropriate title
names(averageData1) = c("Day","Type", "Interval","Average.Steps")

## Graph the time plot using lattice and the layout option
library(lattice)
plot <-xyplot(Average.Steps~Interval|Type, averageData1, type = "l", xlab = "Interval", ylab = "Average Steps",layout = c(1,2))
print(plot)
```






